Dominance Regions for Rank Two Cluster Algebras

### 2022-04-05 dominance_regions.tex
Section 3 edits

\section{Rank Two Lattice Mutations}
  We begin by studying transformations of $\RR^2$ which determine the transformation of $\bfg$-vectors under mutations.

  Write $\phi_0:\RR^2\to\RR^2$ for the identity map and define piecewise-linear maps $\phi_{\pm 1},\phi_{\pm 2}:\RR^2\to\RR^2$ as follows:
  \begin{equation}
    \phi_1(\lambda)
    =
    \begin{cases} 
      (-\lambda_1,c\lambda_1+\lambda_2) & \text{if $\lambda_1 \ge 0$;}\\
      (-\lambda_1,\lambda_2) & \text{if $\lambda_1 < 0$;}
    \end{cases}
    \qquad
    \phi_{-1}(\lambda)
    =
    \begin{cases} 
      (\lambda_1,-\lambda_2) & \text{if $\lambda_2 \ge 0$;}\\
      (\lambda_1+b\lambda_2,-\lambda_2) & \text{if $\lambda_2 < 0$;}
    \end{cases}
  \end{equation}
  \begin{equation}
    \label{eq:forward two step mutation}
    \phi_2(\lambda)
    =
    \begin{cases}
      \big((bc-1)\lambda_1+b\lambda_2, -c\lambda_1-\lambda_2\big) & \text{if $\lambda_1\ge 0$ and $c\lambda_1+\lambda_2\ge 0$;}\\
      (-\lambda_1, -c\lambda_1-\lambda_2) & \text{if $\lambda_1\ge 0$ and $c\lambda_1+\lambda_2<0$;}\\
      (-\lambda_1+b\lambda_2, -\lambda_2) & \text{if $\lambda_1<0$ and $\lambda_2\ge 0$;}\\
      (-\lambda_1,-\lambda_2) & \text{if $\lambda_1<0$ and $\lambda_2<0$;}
    \end{cases}
  \end{equation}
  \begin{equation}
    \label{eq:backward two step mutation}
    \phi_{-2}(\lambda)
    =
    \begin{cases}
      \big(-\lambda_1-b\lambda_2, c\lambda_1+(bc-1)\lambda_2\big) & \text{if $\lambda_2\le 0$ and $\lambda_1+b\lambda_2\le 0$;}\\
      (-\lambda_1-b\lambda_2, -\lambda_2) & \text{if $\lambda_2\le 0$ and $\lambda_1+b\lambda_2>0$;}\\
      (-\lambda_1, c\lambda_1-\lambda_2) & \text{if $\lambda_2>0$ and $\lambda_1\le 0$;}\\
      (-\lambda_1,-\lambda_2) & \text{if $\lambda_2>0$ and $\lambda_1>0$.}
    \end{cases}
  \end{equation}

  For $|k|>2$, define piecewise-linear maps
  \[\phi_k
    :=
    \begin{cases}
      \phi_1\phi_2^j & \text{if $k>0$ and $k=2j+1$;}\\
      \phi_{-1}\phi_{-2}^j & \text{if $k<0$ and $k=-2j-1$;}\\
      \phi_2^j & \text{if $k>0$ and $k=2j$;}\\
      \phi_{-2}^j & \text{if $k<0$ and $k=-2j$;}\\
    \end{cases}
  \]
  and observe that $\phi_{2j}^{-1}=\phi_{-2j}$ for $j\in\ZZ$.

  \begin{definition}
    For $\lambda\in\ZZ^2$, define cones $\cC(\lambda):=\lambda-b\RR_{\ge0}+c\RR_{\ge0}$ and $\cC'(\lambda):=\lambda+b\RR_{\ge0}-c\RR_{\ge0}$.
    The \emph{dominance region} $\cP(\lambda)$ is the intersection of $\bigcap_{j\in\ZZ}\phi_{2j}^{-1}\cC(\phi_{2j}\lambda)$ and $\bigcap_{j\in\ZZ} \phi_{2j+1}^{-1}\cC'(\phi_{2j+1}\lambda)$
  \end{definition}

@@

  Write $\cI \subset \RR^2$ for the \emph{imaginary cone} (positively) spanned by the vectors $\big(2b,-bc\pm\sqrt{bc(bc-4)}\big)$.
  Also, write $\cI' \subset \RR^2$ for the \emph{dual imaginary cone} (positively) spanned by the vectors $\big(-bc\pm\sqrt{bc(bc-4)},2c\big)$.
  %\begin{lemma}
  %  The vectors $\big(2b, -bc+\sqrt{bc(bc-4)}\big)$ (resp. $\big(2b, -bc-\sqrt{bc(bc-4)}\big)$) and $\big(2c, bc+\sqrt{bc(bc-4)}\big)$ (resp. $\big(2c, bc-\sqrt{bc(bc-4)}\big)$) are perpendicular.
  %\end{lemma}
  %\begin{proof}
  %  This is immediate from the equality $\big(-bc+\sqrt{bc(bc-4)}\big)\big(bc+\sqrt{bc(bc-4)}\big)=-4bc$.
  %\end{proof}
  \begin{lemma}
    \label{le:imaginary stability}
    For $j\in\ZZ$ and $\lambda\in\cI$, we have $\phi_{2j}(\lambda)\in\cI$ and $\phi_{2j+1}(\lambda)\in\cI'$.
  \end{lemma}
  \begin{proof}
    Since $\phi_{2j}=\phi_2^j$ and $\phi_{2j+1}=\phi_1\phi_{t_2}^j$, the result follows from the cases $j=1$ and $j=0$, respectively.
    The claim is immediate for $\phi_2$ since the vectors $\big(2b,-bc\pm\sqrt{bc(bc-4)}\big)$ are eigenvectors and $\phi_2$ is linear in $\cI$.
    Now, since $b>0$, we have $\phi_1(\lambda)=(-\lambda_1,c\lambda_1+\lambda_2)$ and so $\phi_1\Big( \big(2b,-bc\pm\sqrt{bc(bc-4)}\big) \Big)=\big(-2b,bc\pm\sqrt{bc(bc-4)}\big)$, which is the $\frac{bc\pm\sqrt{bc(bc-4)}}{2c}$-multiple of the boundary vector for $\cI'$ and the result follows for arbitrary $\lambda\in\cI$ by linearity of $\phi_1$ in this region.
  \end{proof}

  It will be useful to have explicit expressions for $\phi_k(\lambda)$ for $\lambda\in\cI$.
  \begin{lemma}
    \label{le:imaginary transformations}
    For $j\in\ZZ$ and $\lambda\in\cI$, we have
    \begin{align*}
      \phi_{2j}(\lambda)&=(u_{2j+1,+}\lambda_1+u_{2j,+}\lambda_2,-u_{2j,-}\lambda_1-u_{2j-1,-}\lambda_2);\\
      \phi_{2j+1}(\lambda)&=(-u_{2j+1,+}\lambda_1-u_{2j,+}\lambda_2,u_{2j+2,-}\lambda_1+u_{2j+1,-}\lambda_2).
    \end{align*}
  \end{lemma}
  \begin{proof}
    We work by induction on $j$, the case $j=0$ being clear from the definitions.
    For $\lambda\in\cI$, the action of $\phi_2$ from \eqref{eq:forward two step mutation} can be rewritten as
    \[\phi_2(\lambda)=(u_{3,-}\lambda_1+u_{2,+}\lambda_2,-u_{2,-}\lambda_1-u_{1,+}\lambda_2),\]
    and so
    \begin{align*}
      \phi_{2j+2}(\lambda)&=\phi_2\phi_{2j}(\lambda)\\
      &=\big( (u_{3,-}u_{2j+1,+}-u_{2,+}u_{2j,-})\lambda_1+(u_{3,-}u_{2j,+}-u_{2,+}u_{2j-1,-})\lambda_2,\\
      &\qquad -(u_{2,-}u_{2j+1,+}-u_{1,+}u_{2j,-})\lambda_1-(u_{2,-}u_{2j,+}-u_{1,+}u_{2j-1,-})\lambda_2 \big)\\
      &=(u_{2j+3,+}\lambda_1+u_{2j+2,+}\lambda_2, -u_{2j+2,-}\lambda_1-u_{2j+1,-}\lambda_2),
    \end{align*}
    as desired, where the last equality uses \eqref{eq:long Chebyshev recursion} with $i=2j+1$ and $\ell=2$.

    Similarly, using $\phi_1(\lambda)=(-\lambda_1,u_{2,-}\lambda_1+u_{1,+}\lambda_2)$ for $\lambda\in\cI$, together with the basic Chebyshev recursion and the equality $\phi_{2j+1}=\phi_1\phi_2^j$ gives the claimed formula for $\phi_{2j+1}$ from that of $\phi_{2j}$.
  \end{proof}

>>>>

\section{$\bfg$-vector Mutations}
  We begin by studying transformations of $\RR^2$ which determine the change of $\bfg$-vectors under mutations.
  Write $\phi_0:\RR^2\to\RR^2$ for the identity map and define piecewise-linear maps $\phi_{\pm 1}:\RR^2\to\RR^2$ as follows:
  \begin{equation}
    \phi_1(\lambda)
    =
    \begin{cases} 
      (-\lambda_1,c\lambda_1+\lambda_2) & \text{if $\lambda_1 \ge 0$;}\\
      (-\lambda_1,\lambda_2) & \text{if $\lambda_1 < 0$;}
    \end{cases}
    \qquad
    \phi_{-1}(\lambda)
    =
    \begin{cases} 
      (\lambda_1,-\lambda_2) & \text{if $\lambda_2 \ge 0$;}\\
      (\lambda_1+b\lambda_2,-\lambda_2) & \text{if $\lambda_2 < 0$.}
    \end{cases}
  \end{equation}
  For $k\in\ZZ$ with $|k|>1$, define piecewise-linear maps
  \[\phi_k
    :=
    \begin{cases}
      (\phi_{-1}^{-1}\phi_1)^j & \text{if $k=2j$;}\\
      \phi_1(\phi_{-1}^{-1}\phi_1)^j & \text{if $k=2j+1$.}\\
    \end{cases}
  \]
  These determine the dominance region as follows.
  \begin{definition}
    For $\lambda\in\ZZ^2$, define cones 
    \[\cC(\lambda):=\{(\lambda_1-r,\lambda_2+s):r,s\in\RR_{\ge0}\}\quad\text{ and }\quad \cC'(\lambda):=\{(\lambda_1+r,\lambda_2-s):r,s\in\RR_{\ge0}\}.\]
    The \emph{dominance region} $\cP(\lambda)$ is the intersection of $\bigcap_{j\in\ZZ}\phi_{2j}^{-1}\cC(\phi_{2j}\lambda)$ and $\bigcap_{j\in\ZZ} \phi_{2j+1}^{-1}\cC'(\phi_{2j+1}\lambda)$.
  \end{definition}

  We record here a few useful calculations relating to the tropical transformations $\phi_k$.
  First observe the following explicit expression for $\phi_2$:
  \begin{equation}
    \label{eq:forward two step mutation}
    \phi_2(\lambda)
    =
    \begin{cases}
      \big((bc-1)\lambda_1+b\lambda_2, -c\lambda_1-\lambda_2\big) & \text{if $\lambda_1\ge 0$ and $c\lambda_1+\lambda_2\ge 0$;}\\
      (-\lambda_1, -c\lambda_1-\lambda_2) & \text{if $\lambda_1\ge 0$ and $c\lambda_1+\lambda_2<0$;}\\
      (-\lambda_1+b\lambda_2, -\lambda_2) & \text{if $\lambda_1<0$ and $\lambda_2\ge 0$;}\\
      (-\lambda_1,-\lambda_2) & \text{if $\lambda_1<0$ and $\lambda_2<0$;}
    \end{cases}
  \end{equation}
  %\begin{equation}
  %  \label{eq:backward two step mutation}
  %  \phi_{-2}(\lambda)
  %  =
  %  \begin{cases}
  %    \big(-\lambda_1-b\lambda_2, c\lambda_1+(bc-1)\lambda_2\big) & \text{if $\lambda_2\le 0$ and $\lambda_1+b\lambda_2\le 0$;}\\
  %    (-\lambda_1-b\lambda_2, -\lambda_2) & \text{if $\lambda_2\le 0$ and $\lambda_1+b\lambda_2>0$;}\\
  %    (-\lambda_1, c\lambda_1-\lambda_2) & \text{if $\lambda_2>0$ and $\lambda_1\le 0$;}\\
  %    (-\lambda_1,-\lambda_2) & \text{if $\lambda_2>0$ and $\lambda_1>0$.}
  %  \end{cases}
  %\end{equation}

@@

  Write $\cI \subset \RR^2$ for the \emph{imaginary cone} (positively) spanned by the vectors $\big(2b,-bc\pm\sqrt{bc(bc-4)}\big)$.
  Also, write $\cI' \subset \RR^2$ for the \emph{dual imaginary cone} (positively) spanned by the vectors $\big(-bc\pm\sqrt{bc(bc-4)},2c\big)$.
  %\begin{lemma}
  %  The vectors $\big(2b, -bc+\sqrt{bc(bc-4)}\big)$ (resp. $\big(2b, -bc-\sqrt{bc(bc-4)}\big)$) and $\big(2c, bc+\sqrt{bc(bc-4)}\big)$ (resp. $\big(2c, bc-\sqrt{bc(bc-4)}\big)$) are perpendicular.
  %\end{lemma}
  %\begin{proof}
  %  This is immediate from the equality $\big(-bc+\sqrt{bc(bc-4)}\big)\big(bc+\sqrt{bc(bc-4)}\big)=-4bc$.
  %\end{proof}
  \begin{lemma}
    \label{le:imaginary stability}
    For $j\in\ZZ$ and $\lambda\in\cI$, we have $\phi_{2j}(\lambda)\in\cI$ and $\phi_{2j+1}(\lambda)\in\cI'$.
  \end{lemma}
  \begin{proof}
    Since $\phi_{2j}=\phi_2^j$ and $\phi_{2j+1}=\phi_1\phi_{t_2}^j$, the result follows from the cases $j=1$ and $j=0$, respectively.
    The claim is immediate for $\phi_2$ since the vectors $\big(2b,-bc\pm\sqrt{bc(bc-4)}\big)$ are eigenvectors and $\phi_2$ is linear in $\cI$.
    Now, since $b>0$, we have $\phi_1(\lambda)=(-\lambda_1,c\lambda_1+\lambda_2)$.
    It follows that $\phi_1\Big( \big(2b,-bc\pm\sqrt{bc(bc-4)}\big) \Big)=\big(-2b,bc\pm\sqrt{bc(bc-4)}\big)$, which is the $\frac{bc\pm\sqrt{bc(bc-4)}}{2c}$-multiple of the boundary vector for $\cI'$.
    The result then follows for arbitrary $\lambda\in\cI$ by linearity of $\phi_1$ in this region.
  \end{proof}

  It will be useful to have explicit expressions for $\phi_k(\lambda)$ for $\lambda\in\cI$.
  \begin{lemma}
    \label{le:imaginary transformations}
    For $j\in\ZZ$ and $\lambda\in\cI$, we have
    \begin{align*}
      \phi_{2j}(\lambda)&=(u_{2j+1,+}\lambda_1+u_{2j,+}\lambda_2,-u_{2j,-}\lambda_1-u_{2j-1,-}\lambda_2);\\
      \phi_{2j+1}(\lambda)&=(-u_{2j+1,+}\lambda_1-u_{2j,+}\lambda_2,u_{2j+2,-}\lambda_1+u_{2j+1,-}\lambda_2).
    \end{align*}
  \end{lemma}
  \begin{proof}
    We work by induction on $j$, the case $j=0$ being clear from the definitions.
    For $\lambda\in\cI$, the action of $\phi_2$ from \eqref{eq:forward two step mutation} can be rewritten as
    \[\phi_2(\lambda)=(u_{3,-}\lambda_1+u_{2,+}\lambda_2,-u_{2,-}\lambda_1-u_{1,+}\lambda_2).\]
    Therefore
    \begin{align*}
      \phi_{2j+2}(\lambda)&=\phi_2\phi_{2j}(\lambda)\\
      &=\big( (u_{3,-}u_{2j+1,+}-u_{2,+}u_{2j,-})\lambda_1+(u_{3,-}u_{2j,+}-u_{2,+}u_{2j-1,-})\lambda_2,\\
      &\qquad -(u_{2,-}u_{2j+1,+}-u_{1,+}u_{2j,-})\lambda_1-(u_{2,-}u_{2j,+}-u_{1,+}u_{2j-1,-})\lambda_2 \big)\\
      &=(u_{2j+3,+}\lambda_1+u_{2j+2,+}\lambda_2, -u_{2j+2,-}\lambda_1-u_{2j+1,-}\lambda_2),
    \end{align*}
    as desired, where the last equality uses \eqref{eq:long Chebyshev recursion} with $i=2j+1$ and $\ell=2$.

    Similarly, using $\phi_1(\lambda)=(-\lambda_1,u_{2,-}\lambda_1+u_{1,+}\lambda_2)$ for $\lambda\in\cI$ together with the basic Chebyshev recursion and the equality $\phi_{2j+1}=\phi_1\phi_2^j$ gives the claimed formula for $\phi_{2j+1}$ from that of $\phi_{2j}$.
  \end{proof}
